---
title: "INFSCI 2595 Spring 2023 - MIDTERM"
subtitle: "Assigned February 23, 2023; Due: March 2, 2023"
author: "Xia Xiaoyu"
date: "Submission time: March 2, 2023 at 11:00PM EST"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Collaborators

You are **NOT** allowed to collaborate within anyone. Collaboration, copying, and/or cheating of any kind will not be tolerated.  

By downloading the template file you assert that you will **not** collaborate with anyone else on this exam.  

## Overview

This midterm tests your understanding of the concepts, math, and programming required to learn distributions from data. You are required to perform a mixture of derivations and programming to solve the questions on the exam.  

**READ the problem statements CAREFULLY!**  

**IMPORTANT**: The RMarkdown assumes you have downloaded two data sets (CSV files) and saved them to the same directory you saved the template Rmarkdown file. If you do not have the CSV files in the correct location, the data will not be loaded correctly.  

### IMPORTANT!!!

Certain code chunks are created for you. Each code chunk has `eval=FALSE` set in the chunk options. You **MUST** change it to be `eval=TRUE` in order for the code chunks to be evaluated when rendering the document.  

You are free to add more code chunks if you would like.  

## Load packages

This assignment will use packages from the `tidyverse` suite.  

```{r, load_packages}
library(tidyverse)
```

## Problem 01

You have fit discrete and continuous distributions to data, using non-Bayesian and Bayesian approaches. Bayesian analyses require a prior to be formulated, and it can be difficult to understand how a prior is specified in a general setting. This exam seeks to give you some practice doing that by using the **Empirical Bayes** approach. Empirical Bayes is a rather odd sounding name, but the idea is that you will estimate the parameters of the prior using all of the available data. It is useful when the data can be structured into **groups**. Some groups might have many observations, while others may have a limited number of samples. Empirical Bayes is useful when there are many groups (potentially in the thousands) that can be used to estimate the prior parameters. Once estimated, the prior is applied to each group separately. In this manner you have made use of data to understand the relevant bounds on your unknowns and specified those bounds within a prior probability distribution. The *informative* prior is updated based on each group's data to yield the updated belief (the posterior) for each group. (Note that if we would have very few groups we could not use Empirical Bayes and thus would need to use full Bayesian approaches via multilevel, hierarchical, or partial pooling models.)  

To see how the Empirical Bayes process works you will work with a Sports related application. You are interested in learning the catch probability (or catch rate) in the National Football League (NFL). The catch rate is defined as the number of successful receptions (catches) by a player divided by the number of targets (a target corresponds to a pass thrown at the player). You can therefore consider successfully catching a pass as the **event**, and the number of times the player was targeted as the number of **trials**. The probability of catching a pass is therefore the **event probability** we are interested in learning.  

Let's consider you are working on this application because you were recently hired as a sports analytics intern for an NFL team. You are provided with 3 seasons worth of data (2018, 2019, and 2020) of every player with at least 1 target (thus at least 1 trial). Calculating the catch rate is simple to do. It is also easy to search for and find. For example, [here](https://www.pro-football-reference.com/years/2019/receiving.htm) are the catch rates for all NFL players in the 2019 season. You were hired because the NFL team wishes to move away from simple *point estimates*. The team wants to have a better understanding of the **uncertainty** in the performance. Understanding the uncertainty is critical when evaluating talent, and making decisions for which players to sign in free agency.  

You will work with two datasets for this exam. Both are loaded for you in code chunk below. The first, `df_all`, is the larger of the two. The second, `df_focus`, is a subset of `df_all` so that we way can focus on 23 players to help with visualization and discussion in the exam.  

```{r, load_datasets}
file_with_all <- "midterm_all_data.csv"
df_all <- readr::read_csv(file_with_all, col_names = TRUE)

file_with_focus <- "midterm_focus_data.csv"
df_focus <- readr::read_csv(file_with_focus, col_names = TRUE)
```


Both data sets consist of 3 variables, `player_id`, `num_events`, and `num_trials`. The `num_events` is the number of receptions, and `num_trials` is the number of targets. The variables are written in the general terms that we have used in the class rather than NFL specific terms. The `player_id` variable is an ID variable for each player. Thus, one row in either data set tells us the number of receptions and number of targets associated with an individual player over the three seasons. The data in this exam are real and were downloaded from the `nflfastR` package (documentation available [here](https://www.nflfastr.com/index.html) if you are interested). The `player_id` variable is an anonymous identification number I created so that NFL fans cannot easily identify players.  

### 1a)

You will eventually use the Empirical Bayes approach in this exam. However, to help motiviate why such an approach is useful you will begin by assuming you are not capable of creating an informative prior. For example, even if you watch every Pittsburgh Steelers' game, you might not know what the average catch rate is in the NFL. Since you do not feel comfortable specifying reasonable bounds, you decide to use a vague and uninformative prior formulation.  

You will use a Binomial likelihood and a conjugate Beta prior on the unknown catch rate (or event probability in general terms), $\mu$. For generality, you will denote each player with a subscript $j$ and the total number of players as $J$. Thus, the unknown event probability for the $j$-th player is $\mu_j$ where $j=1,...,J$. The posterior distribution on the $j$-th player's unknown catch rate, $\mu_j$ given the $m_j$ catches (events) out of $N_j$ targets (trials) is proportional to:  

$$ 
p\left(\mu_j \mid \left( m, N \right)_j \right) \propto \mathrm{Binomial}\left(m_j \mid \mu_j, N_j\right) \times \mathrm{Beta} \left( \mu_j \mid a, b\right)
$$

Notice that in the above posterior formulation, each player has a potentially distinct event probability, $\mu_j$. The prior consists of two shape hyperparameters, $a$ and $b$. The **SAME** prior shape parameters are applied to every player.  

**You will assume prior shape parameters of $a=0.5$ and $b=0.5$. How many "prior trials" or "prior targets" does this specification correspond to? Why do you think it represents being "uninformed" about the process?**  

#### SOLUTION

What do you think?  

The prior shape parameters correspond to one "prior trial" or "prior target" for each player.
This specification can represent being "uninformed" about the process because it assigns equal weight to all the possible outcomes.The prior parameter does not give any bias to the player's performance results, and a prior with a mean of 0.5 is often used as a non-informative prior in Bayesian analysis without biasing the posterior estimate. 

### 1b)

You are using a conjugate prior to the Binomial likelihood, for each player.  

**What type of distribution is the posterior for the unknown event probability, $\mu_j$, for each player, $j=1,...,J$?**  

#### SOLUTION

What do you think?  

The posterior distribution for the unknown event probability $\mu_j$ for each player $j=1,...,J$ is a beta distribution, which is the conjugate prior of the binomial likelihood.

### 1c)

**Write out the formula for the updated or posterior shape parameters, $a_{new,j}$ and $b_{new,j}$, based on each player's observed number of catches $m_j$ and observed number of targets $N_j$, as well as the prior shape parameters, $a$ and $b$.**  

You do not have to derive the formula for these updated parameters. You may simply write their formula below.  

#### SOLUTION

$$
a_{new,j}=a+m_j
$$
$$
b_{new,j}=b+\left(N_j-m_j \right)
$$

### 1d)

**What is the mean, 0.05 Quantile, 0.95 Quantile, and middle 90% uncertainty interval on the catch rate (event probability) according to the assumed "uninformed" prior?**  

#### SOLUTION
```{r}
a=0.5
b=0.5
mean_1 = a / (a + b)
q05 = qbeta(0.05, a, b)
q95 = qbeta(0.95, a, b)
q90=qbeta(c(0.05,0.95), a, b)
mean_1
q05
q95
q90

```
### 1e)

**Based on your formula in Problem 1c), calculate the updated shape parameters for the `r nrow(df_focus)` players in the `df_focus` `tibble`. You should add two columns using `mutate()` named `anew` and `bnew`. Assign your result to the `post_df_focus_from_vague` object.**  

#### SOLUTION

```{r, solution_01e, eval=TRUE}
post_df_focus_from_vague <-df_focus %>% 
  mutate(anew = 0.5 + num_events,
         bnew = 0.5 + (num_trials - num_events)) 
post_df_focus_from_vague
```


### 1f)

**Calculate the Posterior Mean, Posterior 0.05 Quantile, and the Posterior 0.95 Quantile for each player in `post_df_focus_from_vague`. You should add 3 columns using `mutate()` named `post_avg`, `post_q05`, and `post_q95`. Assign the result to the variable `summary_post_df_focus_from_vague`.**  

#### SOLUTION

```{r, solution_01f, eval=TRUE}
summary_post_df_focus_from_vague <- post_df_focus_from_vague %>% 
  mutate(post_avg = anew / (anew + bnew),
         post_q05 = qbeta(0.05, anew, bnew),
         post_q95 = qbeta(0.95, anew, bnew))

summary_post_df_focus_from_vague
```


### 1g)

You will now visualize the posterior summaries for the `r nrow(df_focus)` players associated with the `df_focus` data set. The bold face font below provides specific instructions for visualizing the posterior summaries for this problem. Please read the instructions carefully! That said, you are free to set colors as you wish.  

**Pipe `summary_post_df_focus_from_vague` into `ggplot()` and map the `x` aesthetic to `as.factor(player_id)`. You will use the `geom_linerange()` to represent the posterior uncertainty by setting the `ymin` and `ymax` aesthetics to `post_q05` and `post_q95`, respectively. You will display the posterior mean with a `geom_point()` by setting the `y` aesthetic to `post_avg`.**  

**Include the maximum likelihood estimate (MLE) on the event probability as an additional `geom_point()` geom by mapping the `y` aesthetic to the correct value, which you must calculate.**  

**Are there players with MLEs that are outside the posterior uncertainty interval? Are there players with posterior mean values that are quite close to the MLEs?**  

#### SOLUTION

```{r, solution_01g}
summary_post_df_focus_from_vague %>% 
  ggplot(mapping = aes(x = as.factor(player_id))) +
  geom_linerange(mapping = aes(ymin = post_q05, ymax = post_q95)) +
  geom_point(mapping = aes(y = post_avg),color="pink") +
  geom_point(mapping = aes(y = num_events / num_trials),color="gray") +
  labs(y = "post_avg & MLE") 
```

What do you think?  

Players with MLE at 0 or 1 are outside the posterior uncertainty interval. There are players whose posterior mean is very close to MLE, such as players with MLE close to 0.5.
### 1h)

**You will create a similar visualization to that from Problem 1g), except instead of mapping the `x` aesthetic to `as.factor(player_id)` you will map the `x` aesthetic to `as.factor(num_trials)`. You must also map the `group` aesthetic in each geom to the `player_id` variable. Doing so allows you to "dodge" the posterior summaries for each player associated with each `num_trials` value.**  

To properly apply the dodging, set the `position` argument to be `position = position_dodge(0.2)` in `geom_linerange()` and both `geom_point()` calls. You should not place `position` inside `aes()`, it should be outside `aes()`.  

**Based on your visualization, which players have high posterior uncertainty on the event probability?**  

#### SOLUTION

```{r, solution_01h}
summary_post_df_focus_from_vague %>% 
  ggplot(mapping = aes(x = as.factor(num_trials))) +
  geom_linerange(mapping = aes(ymin = post_q05, ymax = post_q95,
                               group = player_id),
                 position = position_dodge(0.2)) +
  geom_point(mapping = aes(y = post_avg,
                           color = "pink",
                           group = player_id),
             position = position_dodge(0.2)) +
  geom_point(mapping = aes(y = num_events / num_trials,
                           color = "gray",
                           group = player_id),
             position = position_dodge(0.2)) +
             labs(y = "post_avg & MLE") 
```

What do you think?  
It can be seen from the figure that the more the number of trials, the smaller the posterior uncertainty interval.players with the smallest num_trials have high posterior uncertainty on the event probability.

### 1i)

You will now calculate the posteriors for **ALL** players based on the uninformed prior. Thus, you will not just the limited number of players in the "focused" data set.  

**Calculate the updated shape parameters for all players in the `df_all` `tibble`. You should add two columns using `mutate()` named `anew` and `bnew`. Assign your result to the `post_df_all_from_vague` object.**  

#### SOLUTION
```{r}
post_df_all_from_vague<- df_all%>% 
  mutate(anew = 0.5 + num_events,
         bnew = 0.5 + (num_trials - num_events)) 
post_df_all_from_vague
```


### 1j)

**Calculate the Posterior Mean, 0.05 Quantile, and 0.95 Quantile for each player in `post_df_all_from_vague`. You should add 3 columns using `mutate()` named `post_avg`, `post_q05`, and `post_q95`. Assign the result to the variable `summary_post_df_all_from_vague`.**  

#### SOLUTION

```{r}
summary_post_df_all_from_vague <- post_df_all_from_vague %>% 
  mutate(post_avg = anew / (anew + bnew),
         post_q05 = qbeta(0.05, anew, bnew),
         post_q95 = qbeta(0.95, anew, bnew))

summary_post_df_all_from_vague
```

### 1k)

You will now visualize the Posterior Mean, based on the uniformative prior, relative to the Maximum Likelihood Estimate for the event probability.  

**Create a scatter plot with `ggplot2` where you plot the `post_avg` with respect to the maximum likelihood estimate to the unknown event probability for all players. Map the `color` aesthetic to `num_trials` and include a `geom_abline()` layer with `slope = 1` and `intercept=0`.**  

#### SOLUTION
```{r}
summary_post_df_all_from_vague%>%ggplot()+
  geom_point(mapping = aes(x=num_events / num_trials,y=post_avg,color=num_trials))+
  geom_abline(slope = 1,intercept=0)
```


### 1l)

**Create a scatter plot for the Posterior middle 90% uncertainty interval range (difference between the 0.95 and 0.05 Quantiles) with respect to the `num_trials` using `ggplot2`.**  

#### SOLUTION
```{r}
summary_post_df_all_from_vague%>%ggplot()+
  geom_point(mapping = aes(x=num_trials,y=post_q95-post_q05))+
  geom_abline(slope = 1,intercept=0)
```
  

## Problem 02

In Problem 01, you estimated the unknown event probability for each player separately from all other players. Essentially, you were focused on one player at a time. This style of analysis is known as the **unpooled estimate**, since you are not combining or "pooling" the players (or in general terms the "groups") together.  

The opposite view point is to **completely pool** all players together in order to estimate a single unknown event probability $\mu$. For this, you will assume that all players are independent. Thus the posterior distribution on the unknown "pooled" event probability, $\mu$, is proportional to:  

$$ 
p \left( \mu \mid \left( \left(m, N\right)_j \right)_{j=1}^{J} \right) \propto \prod_{j=1}^{J} \left( \mathrm{Binomial} \left(m_j \mid \mu, N_j \right) \right) \times \mathrm{Beta} \left(\mu \mid a, b\right)
$$

Pay close attention to the subscripts in the above expression. And notice that the prior on the "pooled" unknown $\mu$ relies on the prior shape parameters $a$ and $b$.  

### 2a)

**Write out the log-posterior on the pooled unknown $\mu$ up to a normalizing constant in terms of the observations, $m_j$ and $N_j$ for $j=1,...,J$, and the prior shape parameters, $a$ and $b$. Your result should contain a summation series over the $J$ players.**  

#### SOLUTION
Take the logarithm on both sides to get:
$$
\log\left(p \left( \mu \mid \left( \left(m, N\right)_j \right)_{j=1}^{J} \right)\right) \propto \log\left(\prod_{j=1}^{J} \left( \mathrm{Binomial} \left(m_j \mid \mu, N_j \right) \right) \times \mathrm{Beta} \left(\mu \mid a, b\right)\right)
$$

Simplify the expression on the right to get：
$$
\log\left(\prod_{j=1}^{J} \left( \mathrm{Binomial} \left(m_j \mid \mu, N_j \right)\right)\right)  = \sum_{j=1}^{J} \left( \log \left( \mathrm{Binomial} \left(m_j \mid \mu, N_j \right) \right) \right) =\sum_{j=1}^{J} \left( \log \left( \binom{N_{j}}{m_{j}}\mu^{m_{j}} \left( i - \mu \right)^{N_{j} - m_{j}} \right)  \right)
$$
so the first expression on the right is:
$$
\log\left(\prod_{j=1}^{J} \left( \mathrm{Binomial} \left(m_j \mid \mu, N_j \right)\right)\right) =\log \left( \mu \right) \sum_{j=1}^{J} m_{j} + \log \left( 1 - \mu \right) \sum_{j=1}^{J} \left( N_{j} - m_{j} \right)
$$
As for the second expression:
$$
\log\left(\mathrm{Beta} \left(\mu \mid a, b\right)\right) =\log \left( \frac{\Gamma \left( a + b \right) }{\Gamma \left( a \right) \Gamma \left( b \right)} \mu^{a-1} \left( 1 - \mu \right)^{b-1} \right) = \left( a - 1 \right)  \log \left( \mu \right) + \left( b - 1 \right)  \log \left( 1 - \mu \right)
$$
Finally:
$$
\log\left(p \left( \mu \mid \left( \left(m, N\right)_j \right)_{j=1}^{J} \right)\right) \propto \log \left( \mu \right) \sum_{j=1}^{J} m_{j} + \log \left( 1 - \mu \right) \sum_{j=1}^{J} \left( N_{j} - m_{j} \right) + \left( a - 1 \right)  \log \left( \mu \right) + \left( b - 1 \right)  \log \left( 1 - \mu \right)
$$


### 2b)

The summation series in your solution to 2a) can be simplified by using the average number of events, $\bar{m}$ and the average number of trials $\bar{N}$. The average number of events is defined as:  

$$ 
\bar{m} = \frac{1}{J} \sum_{j=1}^{J} \left( m_j \right)
$$


and the average number of trials is defined as:  

$$ 
\bar{N} = \frac{1}{J} \sum_{j=1}^{J} \left( N_j \right)
$$


**Write your result from 2a) in terms of $\bar{m}$, $\bar{N}$, $J$, and the prior shape parameters $a$ and $b$.**  

#### SOLUTION
From 2a we get:
$$
\log\left(p \left( \mu \mid \left( \left(m, N\right)_j \right)_{j=1}^{J} \right)\right) \propto \log \left( \mu \right) \sum_{j=1}^{J} m_{j} + \log \left( 1 - \mu \right) \sum_{j=1}^{J} \left( N_{j} - m_{j} \right) + \left( a - 1 \right)  \log \left( \mu \right) + \left( b - 1 \right)  \log \left( 1 - \mu \right)
$$

combine with the given formula:
$$
\log\left(p \left( \mu \mid \left( \left(m, N\right)_j \right)_{j=1}^{J} \right)\right) \propto \log \left( \mu \right) \times\bar{m} \times J + \log \left( 1 - \mu \right)  \left( \bar{N}-\bar{m} \right) \times J + \left( a - 1 \right)  \log \left( \mu \right) + \left( b - 1 \right)  \log \left( 1 - \mu \right)
$$


### 2c)

Your expression in 2b) should look familiar.  

**What type of posterior distribution does the unknown "pooled" estimate $\mu$ have?**  

**Write out the formulas for the posterior or updated shape parameters for your specified posterior distribution.**  

#### SOLUTION
The original beta distribution is showed as below:
$$
\mathrm{Beta} \left(\mu \mid a, b\right) \propto \mu^{a-1} \left( 1 - \mu \right)^{b-1}
$$
From 2b we get:
$$
p \left( \mu \mid \left( \left(m, N\right)_j \right)_{j=1}^{J} \right) \propto 
\mu^{\left( J \bar{m} + a - 1 \right)} + \left( 1 - \mu \right)^{\left( J \left( \bar{N} - \bar{m} \right) + b - 1 \right)}
$$
As we compare:
$$
a_{new}=J \bar{m} + a
$$
$$
b_{new}=J \left( \bar{N} - \bar{m} \right) + b
$$

What do you think?  
The type of posterior distribution for the unknown "pooled" estimate $\mu$ have is beta distribution.


### 2d)

**Based on your formula in Problem 2c), calculate the updated shape parameters for the `r nrow(df_focus)` players in the `df_focus` `tibble`. You should add two columns using `mutate()` named `anew` and `bnew`. Assign your result to the `post_df_focus_pooled` object.**  

You will still assume an uninformed prior and thus use $a=b=0.5$ as you did in Problem 01. It is important to remember that we are pooling **ALL** players together to learn the pooled estimate, **NOT** just those in the focused set. You are using the focused set of players right now for visualization purposes.  

#### SOLUTION

```{r, solution_02d, eval=TRUE}
N=count(df_all)$n
num_trials1=mean(df_all$num_trials)
num_events1=mean(df_all$num_events)
post_df_focus_pooled <- df_focus %>% 
  mutate(anew = N * num_events1 + 0.5,
         bnew = N * (num_trials1 - num_events1) + 0.5)
post_df_focus_pooled
```

### 2e)

**Calculate the Posterior Mean, Posterior 0.05 Quantile, and Posterior 0.95 Quantile for each player in `post_df_focus_pooled`. You should add 3 columns using `mutate()` named `post_avg`, `post_q05`, and `post_q95`. Assign the result to the variable `summary_post_df_focus_pooled`.**  

#### SOLUTION

```{r, solution_02e, eval=TRUE}
summary_post_df_focus_pooled <- post_df_focus_pooled %>% 
  mutate(post_avg = anew / (anew + bnew),
         post_q05 = qbeta(0.05, anew, bnew),
         post_q95 = qbeta(0.95, anew, bnew))
summary_post_df_focus_pooled
```

### 2f)

**Pipe `summary_post_df_focus_pooled` into `ggplot()` and map the `x` aesthetic to `as.factor(player_id)`. You will use the `geom_linerange()` to represent the posterior uncertainty by setting the `ymin` and `ymax` aesthetics to `post_q05` and `post_q95` respectively. You will display the posterior mean with a `geom_point()` by setting the `y` aesthetic to `post_avg`. Include the maximum likelihood estimate (MLE) on the event probability as an additional `geom_point()` geom by mapping the `y` aesthetic to the correct value, which you must calculate.**  

**Are there players with MLEs that are outside the posterior uncertainty interval? Are there players with posterior mean values that are quite close to the MLEs?**  

#### SOLUTION

```{r, solution_02f}
summary_post_df_focus_pooled %>% 
  ggplot(mapping = aes(x = as.factor(player_id))) +
  geom_linerange(mapping = aes(ymin = post_q05, ymax = post_q95)) +
  geom_point(mapping = aes(y = post_avg),color="pink") +
  geom_point(mapping = aes(y = num_events / num_trials),color="gray") +
  labs(y = "post_avg & MLE") 
```

What do you think?  
It can be seen from the figure that all players have the same posterior mean, so there is no posterior uncertainty interval. Also, the posterior mean for players with id of 232 and 636 is very close to MLE.

### 2g)

Your visualization in Problem 2f) should not "feel right". Something should seem off.  

**Why does the "pooled" estimate seem incorrect for this application?**  

#### SOLUTION

What do you think?  
It is impossible that all players have the same posterior mean, also there is no posterior uncertainty interval.


## Problem 03

You have now worked through two extremes, the **unpooled** and the completely **pooled** estimates on the unknown event probabilities. You will now try to **blend** the two approaches to reach a compromise by using the Empirical Bayes approach.  

As stated at the beginning of the document, Empirical Bayes estimates the prior from data. In this setting you are interested in deciding informative values for the prior shape parameters, $a$ and $b$, of the Beta prior on each $\mu_j$. If you have a relevant informative prior you will be able to apply that prior to each player separately (the unpooled approach) while "borrowing strength" from the rest of the data. The Empirical Bayes approach is an approximation to more formal *partial pooling* models where groups with larger sample sizes help estimate parameters associated with small sample size groups. Empirical Bayes is useful when there are hundreds to thousands of separate groups. Estimating the prior shape parameters from many groups allows specifying relevant informative priors without requiring numerous conversations with Subject Matter Experts (SMEs) and allows the data to provide representative bounds.  

### 3a)

The Beta prior defines the prior belief on a probability (a fraction or proportion). From an Empirical Bayes approach, you can therefore view the "data" of interest as the observed "catch rate".  

**Plot the histogram of the "catch rate" for all players in the `df_all` data set. Use the `geom_histogram()` geom and set the `binwidth` to be 0.05.**  

#### SOLUTION
```{r}
df_all %>% 
  ggplot(mapping = aes(x = num_events/num_trials)) +
  geom_histogram(binwidth = 0.05) 
```
### 3b)

**Plot the histogram for all "catch rates" in the `df_all` data set again. However, this time use `facet_wrap()` to break up the visualization into `num_trials > 24`.**  

**What can you say about the observations of the players with greater than 25 targets?**  
#### SOLUTION
```{r}
df_all %>% 
  ggplot(mapping = aes(x = num_events/num_trials)) +
  geom_histogram(binwidth = 0.05) +
  facet_wrap( ~ num_trials > 24, 
  nrow = NULL,
  ncol = NULL,
  scales = "fixed",
  shrink = TRUE,
  labeller = "label_value",
  as.table = TRUE,
  drop = TRUE,
  dir = "h",
  strip.position = "top" )
```

What do you think?  
Of the results shown on the left with less than 25 trials, the most notable feature is that there were about 107 players with a "catch rate" of 1. After the sample size is greater than 24 in the right figure, the catch rate is more in line with the beta distribution, and there are no players with a catch rate of 0 and 1, which shows that when the sample size is insufficient, the experimental results are often unreliable.

### 3c)

To keep things simple for this exam, you will estimate the prior shape parameters, $a$ and $b$, based only on the players with greater than 24 targets.  

**Use the `filter()` function to keep all players with greater than 24 targets and assign the result to the `df_24` object. Use the `summary()` function to check the summary stats on `num_trials` to make sure you performed the operation correctly.**  

#### SOLUTION

```{r, solution_03c, eval=TRUE}
df_24 <- df_all %>% filter(num_trials > 24)
df_24
df_24 %>% summary()
```


### 3d)

Since the "catch rate" is a fraction, we can use a Beta distribution as the likelihood of the "fraction" given the shape parameters. Those shape parameters, $a$ and $b$, are unknown and so you must estimate them from the data. Within the Empirical Bayes approach, you will treat this step as finding $a$ and $b$ which **maximize the likelihood**, and so you will not specify prior distributions on the parameters.  

Each observation of the "catch rate" is assumed conditionally independent given the unknown $a$ and $b$ shape parameters. The observed "catch rate" will be denoted as, $\theta_j$, for each player and is defined as:  

$$ 
\theta_j = \frac{m_j}{N_j}
$$

The likelihood on all $j=1,...,J$ catch rates is therefore the product of $J$ conditionally independent Beta distributions:  

$$ 
p \left( \left(\theta_j\right)_{j=1}^{J} \mid a, b\right) = \prod_{j=1}^{J} \mathrm{Beta} \left( \theta_j \mid a, b \right)
$$

**You will define a log-likelihood function in the style of the log-posterior functions we have used so far this semester by completing the two code chunks below.**  

**In the first code chunk, the list of required information, `info_for_ab`, is defined and contains a single variable `theta`. You must calculate it based on the players in the `df_24` data set.**  

**The second code chunk defines the `my_beta_loglik()` function. The first argument, `unknowns`, is the vector of unknown parameters. The second argument, `my_info`, is the list of required information. The comments and variable names provide hints for actions you should perform to calculate the log-likelihood.**  

**The $a$ and $b$ parameters are lower-bounded at zero and thus you must apply the log-transformation to both parameters. You must properly account for the log-derivative adjustment on both parameters when you calculate the log-likelihood.**  

*NOTE*: Several test points are provided for you to check that you have coded your function correctly.  

#### SOLUTION

Define the list of required information. The observed data in your `my_beta_loglik()` must be named `theta`.  

```{r, solution_03d_a, eval=TRUE}
info_for_ab <- list(
  theta = df_24$num_events / df_24$num_trials
)
info_for_ab
```

Define the Beta log-likelihood. The first element in `unknowns` is the log-transformed $a$ parameter and the second element is the log-transformed $b$ parameter. **You are allowed to use built-in density functions to complete this question.**  

```{r, solution_03d_b, eval=TRUE}
my_beta_loglik <- function(unknowns, my_info)
{
  # unpack the log-transformed shape parameters
  log_a <- unknowns[1]
  log_b <- unknowns[2]
  
  # back transform
  a <- exp(log_a)
  b <- exp(log_b)
  
  # calculate the log-likelihood for all observations
  log_lik <-sum(dbeta(my_info$theta, shape1=a, shape2=b, log=TRUE))
  
  # account for the change of variables
  value3<-log_lik + log_a + log_b
  value3
}
```

Try out values of -2 for both log-transformed parameters. If your function is coded correctly you should get a value of -571.8519.  

```{r, solution_03d_c}
value4<-my_beta_loglik(c(-2, -2), info_for_ab)
value4
```

Try out values of 2.5 for both log-transformed parameters. If your function is coded correctly you should get a value of -254.3934.  

```{r, solution_03d_d}
value5<-my_beta_loglik(c(2.5, 2.5), info_for_ab)
value5
```


### 3e)

You will now identify the maximum likelihood estimates for $a$ and $b$. You should use the `optim()` function to manage the optimization for you. Be sure to specify the arguments to `optim()` to make sure that `optim()` knows to *MAXIMIZE* and not *MINIMIZE* the function. Set the `method` argument to `"BFGS"` when you call `optim()`. The gradient argument should be set to `NULL`, `gr=NULL`.  

**Try out two different starting guesses values. The first guess, `init_guess_01`, should be zeros for both parameters and the second guess, `init_guess_02`, should be -1 for both parameters.**  

**Assign your `optim()` results to `log_ab_opt_01` and `log_ab_opt_02`.**  

**Do you get the same parameter estimates regardless of your initial guess?**  

#### SOLUTION

Set the initial guesses.  

```{r, solution_03e, eval=TRUE}
init_guess_01 <- c(0,0)
init_guess_02 <- c(-1,-1)
```

Perform the optimization using the first starting guess.  

```{r, solution_03e_b, eval=TRUE}
log_ab_res_01 <- optim(init_guess_01,
                       my_beta_loglik,
                       gr=NULL,
                       info_for_ab,
                       method="BFGS",
                       lower = -Inf, upper = Inf,
                       control = list(fnscale=-1),
                       hessian=TRUE)
log_ab_res_01
```

Perform the optimization using the second starting guess.  

```{r, solution_03e_c, eval=TRUE}
log_ab_res_02 <- optim(init_guess_02,
                       my_beta_loglik,
                       gr=NULL,
                       info_for_ab,
                       method="BFGS",
                       lower = -Inf, upper = Inf,
                       control = list(fnscale=-1),
                       hessian=TRUE)
log_ab_res_02
```

**Are the identified log-transformed estimates the same?**  
The identified log-transformed estimates are the same.

### 3f)

The optimal parameters in the Problem 3e) are in the log-transformed space.  

**You must back-transform them to calculate the estimates for the prior $a$ and $b$ shape hyperparameters. Assign the back-transformed parameters to `ab_emp_bayes`.**  

**How many a-priori trials does your estimated hyperparameters represent?**  

#### SOLUTION

```{r, solution_03f, eval=TRUE}
ab_emp_bayes <- exp(log_ab_res_01$par)
ab_emp_bayes
```

How many a-priori trials?  
There are about 16 a-priori trials.


### 3g)

You will now visualize the prior distribution you calculated using the Empirical Bayes approach and compare it to the histogram of the observed "catch rates" for all players with more than 24 targets.  

**Complete the two code chunks below. In the first, set the `x` variable within the `prior_for_viz` `tibble` to be 1001 evenly spaced points between the minimum observed catch rate in `df_24` and the maximum observed catch rate in `df_24`. Pipe the result into `mutate()` and calculate the beta density using the `ab_emp_bayes` shape parameters and assign the result to the `beta_pdf` variable.**  

**In the second code chunk, pipe the `df_24` `tibble` into `ggplot()` and map the `x` aesthetic to the observed catch rates. Use a `geom_histogram()` geom and set the `binwidth` to be 0.05. Modify the `y` aesthetic so that way `geom_histogram()` displays the estimated density on the `y` axis instead of the count. To do so you must set `y=stat(density)` within `aes()`. Include a `geom_line()` geom and specify the `data` argument to be the `prior_for_viz` object and map the `x` and `y` aesthetics to `x` and `beta_pdf`, respectively. Set the `color` argument (outside the `aes()` call) to be `'red'` and the `size` argument to 1.15.**  

**How does the empirically derived prior distribution on the event probability compare to the observed histogram of the catch rates?**  

**IMPORTANT**: If you are *not* comfortable with your `ab_emp_bayes` values, you may use `shape1=13` and `shape2=8`. These are **not** the correct answers, though they are in the right ballpark...  

#### SOLUTION

Calculate the Beta PDF based on the calculated prior hyperparameters.  

```{r, solution_03g, eval=TRUE}
prior_for_viz <- tibble::tibble(
  x = seq(min(info_for_ab$theta), max(info_for_ab$theta), length.out = 1001)
) %>% 
  mutate(beta_pdf = dbeta(x, ab_emp_bayes[1], ab_emp_bayes[2]))
```

Visualize the derived prior relative to the observed "catch rates" in the data set.  

```{r, solution_03g_b}
df_24 %>% 
  ggplot(mapping = aes(x = num_events / num_trials)) +
  geom_histogram(binwidth = 0.05,
                 mapping = aes(y = stat(density))) +
  geom_line(data = prior_for_viz,
            mapping = aes(x = x, y = beta_pdf),
            color = "red", size = 1.15) 
```
The empirically derived prior distribution of event probabilities is similar when compared to the observed catch rate histogram.


### 3h)

**Calculate the Mean, 0.05 Quantile, 0.95 Quantile, and middle 90% uncertainty interval associated with your informative prior.**  

*IMPORTANT*: If you are *not* comfortable with your `ab_emp_bayes` values, you may use `shape1=13` and `shape2=8`. These are **not** the correct answers, though they are in the right ballpark...  

#### SOLUTION
Use `shape1=13` and `shape2=8`
```{r}
shape1=13
shape2=8
mean_2 = shape1 / (shape1 +shape2)
q05_1 = qbeta(0.05, shape1, shape2)
q95_1 = qbeta(0.95, shape1, shape2)
q90_1= qbeta(c(0.05,0.95), shape1, shape2)

mean_2 
q05_1 
q95_1
q90_1
```
Use `ab_emp_bayes` values:
```{r}
mean_3 =ab_emp_bayes[1] / (ab_emp_bayes[1] +ab_emp_bayes[2])
q05_2 = qbeta(0.05, ab_emp_bayes[1], ab_emp_bayes[2])
q95_2 = qbeta(0.95, ab_emp_bayes[1], ab_emp_bayes[2])
q90_2=qbeta(c(0.05,0.95), ab_emp_bayes[1], ab_emp_bayes[2])

mean_3
q05_2
q95_2
q90_2
```
### 3i)

**How do the Mean and middle 90% uncertainty intervals compare between the original uninformed prior and the Empirical Bayes derived informative prior?**


#### SOLUTION
```{r}
D_mean=mean_3-mean_1
D_mean
D_90=(q90[2]-q90[1])-(q90_2[2]-q90_2[1])
D_90
```

What do you think?  
They are different.
Mean for the Empirical Bayes derived informative prior is 16% greater than the original uninformed prior.
The middle 90% uncertainty intervals for the  original uninformed prior is about 67% bigger than the Empirical Bayes derived informative prior.



## Problem 04

You now have everything in place to calculate the posterior on the event probability associated with each player, $\mu_j$. The $a$ and $b$ parameters that you had originally set to 0.5, are now equal to your Empirical Bayes estimated values.  

If you are not comfortable with your estimates you may use the same values as in Problem 3g) of `shape1=13` and `shape2=8`.  

### 4a)

**Calculate the updated or new shape parameters for the players in the `df_focus` `tibble`. You should add two columns using `mutate()` named `anew` and `bnew`. Assign your result to the `post_df_focus_empbayes` object.**  

#### SOLUTION

```{r, solution_04a, eval=TRUE}
post_df_focus_empbayes <- df_focus %>% 
  mutate(anew = num_events + ab_emp_bayes[1],
         bnew = (num_trials - num_events) + ab_emp_bayes[2])
post_df_focus_empbayes
```

### 4b)

**Calculate the posterior Mean, Posterior 0.05 Quantile, and Posterior 0.95 Quantile for each player in `post_df_focus_empbayes`. You should add 3 columns using `mutate()` named `post_avg`, `post_q05`, and `post_q95`. Assign the result to the variable `summary_post_df_focus_empbayes`.**  

#### SOLUTION

```{r, solution_04b, eval=TRUE}
summary_post_df_focus_empbayes <- post_df_focus_empbayes %>% 
  mutate(post_avg = anew / (anew + bnew),
         post_q05 = qbeta(0.05, anew, bnew),
         post_q95 = qbeta(0.95, anew, bnew))
```

### 4c)

You will repeat the visualizations from Problem 1) to understand the effect of your informative prior distribution.  

**Pipe `summary_post_df_focus_empbayes` into `ggplot()` and map the `x` aesthetic to `as.factor(player_id)`. You will use the `geom_linerange()` to represent the posterior uncertainty by setting the `ymin` and `ymax` aesthetics to `post_q05` and `post_q95` respectively. You will display the posterior mean with a `geom_point()` by setting the `y` aesthetic to `post_avg`. Include the maximum likelihood estimate (MLE) on the event probability as an additional `geom_point()` geom by mapping the `y` aesthetic to the correct value, which you must calculate.**  

**How does this visualization compare to those you made using the vague unpooled estimate and the completely pooled estimate?**  

#### SOLUTION
```{r}
summary_post_df_focus_empbayes  %>% 
  ggplot(mapping = aes(x = as.factor(player_id))) +
  geom_linerange(mapping = aes(ymin = post_q05, ymax = post_q95)) +
  geom_point(mapping = aes(y = post_avg),color="pink") +
  geom_point(mapping = aes(y = num_events / num_trials),color="gray") +
  labs(y = "post_avg & MLE") 
```
It can be seen from the figure that in this case the posterior mean is more concentrated in the interval from 0.5 to 0.75. And the posterior probability uncertainty interval is smaller.
### 4d)

**You will create a similar visualization, except instead of mapping the `x` aesthetic to `as.factor(player_id)` you will map the `x` aesthetic to `as.factor(num_trials)`. You must also map the `group` aesthetic in each geom to the `player_id` variable. Doing so allows you "dodge" the posterior summaries for each player associated with each `num_trials` value.**  

To properly apply the dodging, set the `position` argument to be `position = position_dodge(0.2)` in `geom_linerange()` and both `geom_point()` calls. You should not place `position` inside `aes()`, it should be outside `aes()`.  

#### SOLUTION
```{r}
summary_post_df_focus_empbayes  %>% 
  ggplot(mapping = aes(x = as.factor(num_trials))) +
  geom_linerange(mapping = aes(ymin = post_q05, ymax = post_q95,
                               group = player_id),
                 position = position_dodge(0.2)) +
  geom_point(mapping = aes(y = post_avg,
                           color = "pink",
                           group = player_id),
             position = position_dodge(0.2)) +
  geom_point(mapping = aes(y = num_events / num_trials,
                           color = "gray",
                           group = player_id),
             position = position_dodge(0.2)) +
             labs(y = "post_avg & MLE") 
```


### 4e)

You will now calculate the posteriors for **ALL** players using the Empirical Bayes approach, not just the limited number of players in the "focused" data set.  

**Calculate the updated shape parameters for all players in the `df_all` `tibble`. You should add two columns using `mutate()` named `anew` and `bnew`. Assign your result to the `post_df_all_empbayes` object.**  

#### SOLUTION

```{r}
post_df_all_empbayes <- df_all %>% 
  mutate(anew = num_events + ab_emp_bayes[1],
         bnew = (num_trials - num_events) + ab_emp_bayes[2])
post_df_all_empbayes
```


### 4f)

**Calculate the Posterior Mean, Posterior 0.05 Quantile, and Posterior 0.95 Quantile for each player in `post_df_all_empbayes`. You should add 3 columns using `mutate()` named `post_avg`, `post_q05`, and `post_q95`. Assign the result to the variable `summary_post_df_all_empbayes`.**  

#### SOLUTION
```{r}
summary_post_df_all_empbayes<- post_df_all_empbayes %>% 
  mutate(post_avg = anew / (anew + bnew),
         post_q05 = qbeta(0.05, anew, bnew),
         post_q95 = qbeta(0.95, anew, bnew))
```


### 4g)

You will now visualize the Posterior Mean, based on the Empirical Bayes informative prior, relative to the Maximum Likelihood Estimate for the event probability.  

**Create a scatter plot with `ggplot2` where you plot the `post_mean` with respect to the maximum likelihood estimate to the unknown event probability for all players. Map the `color` aesthetic to `num_trials` and include a `geom_abline()` layer with `slope = 1` and `intercept=0`.**  

#### SOLUTION

```{r}
summary_post_df_all_empbayes%>%ggplot()+
  geom_point(mapping = aes(x=num_events / num_trials,y=post_avg,color=num_trials))+
  geom_abline(slope = 1,intercept=0)
```


### 4h)

**Create a scatter plot for the Posterior middle 90% uncertainty interval range (difference between the 0.95 and 0.05 Quantiles) with respect to the `num_trials` using `ggplot2`.**  

#### SOLUTION
```{r}
summary_post_df_all_empbayes%>%ggplot()+
  geom_point(mapping = aes(x=num_trials,y=post_q95-post_q05))+
  geom_abline(slope = 1,intercept=0)
```


### 4i)

The code chunk below vertically concatenates (binds) the `summary_post_df_all_from_vague` and `summary_post_df_all_empbayes` `tibble`s into a single `tibble` for you. This new `tibble`, `compare_post_summaries` includes a new column, `from_prior`, which denotes if the prior was the original uninformed prior or the Empirical Bayes informative prior.  

```{r, assemble_both_types_to_compare, eval=TRUE}
compare_post_summaries <- summary_post_df_all_from_vague %>% 
  mutate(from_prior = 'uninformed') %>% 
  bind_rows(summary_post_df_all_empbayes %>% 
              mutate(from_prior = 'Empirical Bayes'))

compare_post_summaries %>% glimpse()
```


You will use this "composite" `tibble` to examine the sensitivity of the Posterior to the prior choice and sample size.  

**Create a scatter plot for the Posterior middle 90% uncertainty interval range (difference between the 0.95 and 0.05 Quantiles) with respect to the `num_trials` using `ggplot2`. This plot will be similar to the one from 4h) except this time you must use the `compare_post_summaries` `tibble` instead of `summary_post_df_all_empbayes`. You must map the `color` aesthetic to the `from_prior` variable within the scatter plot.**  

#### SOLUTION
```{r}
compare_post_summaries%>%ggplot()+
  geom_point(mapping = aes(x=num_trials,y=post_q95-post_q05,color=from_prior))+
  geom_abline(slope = 1,intercept=0)
```


### 4j)

**Based on your visualizations in this exam, discuss how an informative prior influences the Posterior when the sample size is small compared with large sample sizes.**  

#### SOLUTION

What do you think?  

An informative prior can have a greater influence on the posterior distribution when the sample size is small compared to large sample sizes. This is because the prior provides additional information to the model, which can dominate the posterior when there is little data available. However, as the sample size increases, the likelihood function becomes more informative and the influence of the prior decreases, leading to a posterior distribution that is more heavily influenced by the data.

## Problem 05

Now that you have posterior distributions based on an informative prior for every player in the data set, it is time to consider answering a question the NFL team is interested in. The team wants to identify the best receivers in the data set, and it wants to be confident in that selection. Your Bayesian analysis allows answering probabilistic questions. You will answer several such questions now.  

### 5a)

**Calculate the probability that each player has a catch rate (event probability) of greater than 0.67. Add a column to the `summary_post_df_all_empbayes` object named `prob_grt_67`. Assign the result to a new variable `post_player_eval`.**  

#### SOLUTION

```{r}
prob_grt_67= 1 - pbeta(0.67, summary_post_df_all_empbayes$anew, summary_post_df_all_empbayes$bnew)

post_player_eval <- summary_post_df_all_empbayes %>% 
  mutate(prob_grt_67)
post_player_eval
```
### 5b)

**Identify the top 10 players based on the posterior probability that their catch rate is greater than 0.67. What do these players all have in common, besides the `prob_grt_67` value?**  

#### SOLUTION
```{r}
arrange(post_player_eval,desc(prob_grt_67)) %>%  head(10)
```
For these 10 player, whose posterior probability that their catch rate is greater than 0.67 are all close to 1. The posterior mean, 5th quantile, and 95th quantile for each player is similar to each other.


### 5c)

**Identify the 10 players with the lowest posterior probability that their catch rate is greater than 0.67. What is the smallest number of targets (trial size) associated with these 10 players?**  

#### SOLUTION

```{r}
arrange(post_player_eval,prob_grt_67) %>% head(10)
```
The smallest number of targets (trial size) associated with these 10 players is 47.

### 5d)

A player with a large sample size could mean that player is well known, especially within the NFL. The team is interested in identifying players that are not as well known, and yet seem to have high catch rates.  

**Identify 10 players with the smallest sample sizes (number of trials) while still having `prob_grt_67` values greater than 0.75.**  

#### SOLUTION

```{r}
arrange(filter(post_player_eval,prob_grt_67 > 0.75),num_trials) %>% head(10)
```


### 5e)

**Why do you think the questions in this problem were focused on calculating the probability that the catch rate is greater than 0.67? What is the interpretation of such a question?**  

*HINT*: Consider the interpretation of the completely pooled estimate.  

#### SOLUTION

What do you think?  

The purpose of our experiment is to understand the probability of receiving the ball in the National Football League (NFL) as a basis for evaluating talent and deciding which players to sign to free agency. Calculating the probability that the catch rate is greater than 0.67 can help us better select players to sign, because this probability is close to the average level of players in the team.
### 5f)

Sometimes teams are willing to take risks on players that they are uncertain about. Teams hope the player will succeed, but they know there is a chance the player could not meet expectations. To manage risks, these players are signed to contracts with less guaranteed money but offer substantial bonuses should the player meet performance goals to help incentive the player.  

The posterior distribution allows you to quantify the uncertainty on each player. You will measure the uncertainty as the range of the posterior 90% uncertainty interval.  

**Create a scatter plot in `ggplot2` where you plot the posterior 90% uncertainty interval with respect to the event probability's posterior mean. Color the markers by the number of trials.**  

#### SOLUTION

```{r}
post_player_eval %>% 
  ggplot() +
  geom_point(mapping = aes(x=post_avg,y=post_q95-post_q05,color=num_trials))
```


### 5g)

**Which players would you recommend to the team to take a risk on based on your figure in 5f)?**  

#### SOLUTION

What do you think?  
If we want to take risk, we should choose those players who have high post mean and high uncertainty interval, so we'd better choose the player located in the right part on the 5f image.